##Reproducible Research : Peer Assessment 1
##Author : Tom Luijten

###Load the required packages
```{r loadpackages, echo = TRUE, results='hide'}
library(dplyr)
library(ggplot2)
library(lattice)
```

###Loading and preprocessing the data
First of all, let's load the data:
```{r loaddata, echo = TRUE}
activity <- read.csv(file="activity.csv", header=TRUE, sep=",")
```

```{r preprocessing, echo = TRUE}
activity$date <- as.Date(activity$date, "%Y-%m-%d") 
```

### Calculating the mean total number of steps taken per day
Calculate the total number of steps taken per day
```{r totalsteps, echo = TRUE}
grp <- group_by(activity, date)
totalday <- summarise(grp, sum = sum(steps,na.rm = TRUE))
```

Make a histogram of the total number of steps taken each day
```{r histogram, echo = TRUE}
qplot(x = sum, data = totalday, ylim = c(0,10))+
  geom_histogram() +
  labs(x = "Number of steps", y = "Frequency")
```

Calculate the mean and median of the total number of steps taken per day
```{r meanmed, echo = TRUE}
meansum <- mean(totalday$sum, na.rm= TRUE)
mediansum <- median(totalday$sum, na.rm= TRUE)
```
The mean of the total number of steps taken per day is equal to `r meansum` and the median is equal to `r mediansum`.

### Calculate the average daily activity pattern
Calculate average number of steps taken in each 5 minute interval
```{r averagesteps, echo = TRUE}
grp <- group_by(activity, interval)
averageinterval <- summarise(grp, steps = mean(steps,na.rm = TRUE))
plot(x = averageinterval$interval, averageinterval$steps, type = "l", xlab =  "5-minute interval", ylab = "Average number of steps")

averageinterval$interval[which(averageinterval$steps == max(averageinterval$steps))]
```
The 5-minute interval which contains the maximum number of steps across all the days in the dataset, is the `r averageinterval$interval[which(averageinterval$steps == max(averageinterval$steps))]` interval.

### Imputing missing values
Calculate and report the total number of missing values in the dataset
```{r amountmissing, echo = TRUE}
sum(is.na(activity$steps))
```
The total number of missing values in the dataset is equal to `r sum(is.na(activity$steps))`.

Replace missing values by the mean for that 5-minute interval and create the activity2 interval that contains the resplaced values.
```{r replacemissing, echo = TRUE}
activity2 <- activity
for (i in 1:nrow(activity2)) {
  if (is.na(activity2[i,"steps"])) {
    activity2[i,"steps"] <- mean(activity[activity2$interval == activity2[i,"interval"],"steps"],na.rm = TRUE)
  }
}
```

Make a histogram of the total number of steps taken each day
```{r histogram2, echo = TRUE}
grp2 <- group_by(activity2, date)
totalday2 <- summarise(grp2, sum = sum(steps,na.rm = TRUE))
qplot(x = sum, data = totalday2, ylim = c(0,10))+
  geom_histogram() +
  labs(x = "Number of steps", y = "Frequency")
```

Calculate the mean and median of the total number of steps taken per day
```{r meanmed2, echo = TRUE}
meansum2 <- mean(totalday2$sum)
mediansum2 <- median(totalday$sum)
```
The mean of the total number of steps taken per day is equal to `r meansum2` and the median is equal to `r mediansum2`.
From this, it can be concluded that replacing the missing values by means has an effect on the total mean but not on the median. 

### The differences in activity patterns between weekdays and weekends
Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r weekweekend, echo = TRUE}
activity2$weekdayend <- ifelse(weekdays(activity2$date) %in% c("maandag","dinsdag","woensdag","donderdag","vrijdag"), "weekday", "weekend")
```

Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.
```{r panelplot, echo = TRUE}
grp3a <- group_by(activity2[activity2$weekdayend == "weekday",], interval)
averageinterval3a <- cbind(summarise(grp3a, steps = mean(steps,na.rm = TRUE)),"weekday")
names(averageinterval3a) <- c("interval","meansteps", "weekdayend")
grp3b <- group_by(activity2[activity2$weekdayend == "weekend",], interval)
averageinterval3b <- cbind(summarise(grp3b, steps = mean(steps,na.rm = TRUE)),"weekend")
names(averageinterval3b) <- c("interval","meansteps", "weekdayend")
averagewde <- rbind(averageinterval3a,averageinterval3b)

xyplot(averagewde$meansteps ~ averagewde$interval | averagewde$weekdayend, 
       layout = c(1, 2), 
       type = "l", 
       xlab = "Interval", 
       ylab = "Number of steps")


```
